Exploring the Role of Small Molecules in Biological Systems Using Network Approaches

### 2015-02-23 paper.tex
Moved and updated section on using R for network analysis in drug discovery

\newcommand{\rcdk}{\texttt{rcdk}\ }

@@

@@

\section{Handling Small Molecules in R}
\label{sec:handl-small-molec}

Though small molecule are represented in text based formats (SMILES,
SDF, MOL2, InChI, etc.), the R environment does not support parsing
and subsequent manipulation of small molecule structures. The
traditional approach has been to compute numerical features for small
molecules and perform cheminformatics related manipulation outside R
and import the results of such operations into the R workspace. An
alternative approach is to integrate cheminformatics toolkits such as
the CDK \cite{Steinbeck:2003bh}, RDKit or Indigo into the R
environment. Currently there are two packages that support this -
\texttt{ChemmineR} \cite{Cao:2008fj} and \rcdk \cite{Guha:2007aa}. In this
section we briefly review the functionality provided by the \rcdk
package, which is an idiomatic R wrapper around the CDK Java library,
primarily focused on manipulating molecular structures.

Working with small molecule structure data can be broadly grouped into
three tasks - input/output, manipulation \& modification of structures
and computations on molecular structures. The \rcdk package supports
input of all chemical file formats supported by the CDK, including
SMILES, MOL MOL2, SDF and PDB formats. When used with the
\texttt{rinchi}, input and output of the InChI format is also
supported. Chemical structure files can be read locally or over the
internet. In addition to generic file loading, helper functions are
available for the SMILES format, given its ubiquitous use as a
structure exchange format.

Molecular structures are loaded in as references to Java objects. That
is, they are not native R data structures and thus can only be
manipulated using methods from the \texttt{rJava} package. While
inconvenient, the low level details are hidden from view and the \rcdk
package provides an idiomatic R interface to various CDK methods that
operate on molecules, bonds and atoms. For example, given a SMILES
string one can count the number of aromatic atoms using the following
R code
\begin{lstlisting}
mol <- parse.smiles('CNCc1ccccc1')[[1]]
length(which(sapply(get.atoms(mol), 
    function(atom) 
        is.aromatic(atom))))  
\end{lstlisting}
Other methods support operations on bonds, identifying substructures
and retrieving atom or bond properties.

From an analytic point of view, functions that perform computations on
molecular structures are probably the most useful. Such computations
can range from evaluating molecular descriptors \cite{Guha:2012vn} and
generating fingerprints to computing a variety of similarities and so
on. The \rcdk package provides a simple interface to descriptor and
fingerprint calculation. Importantly, the results from these functions
can easily be employed in a network based approach when coupled with
\texttt{igraph}. As an example, consider a similarity network
constructed for 100 molecules, using a Tanimoto cutoff of 0.75

\begin{lstlisting}
library(rcdk)
library(igraph)
library(fingerprint)
mols <- load.molecules('data/mipe-std.smi')
fps <- lapply(mols, get.fingerprint)
smat <- fp.sim.matrix(fps)
smat[ smat < 0.75 ] <- 0
g <- graph.adjacency(smat, mode='undirected', 
    weight=TRUE, diag=FALSE)  
\end{lstlisting}

@@

\section{R as a Platform for Computational Drug Discovery}
\label{sec:r-as-platform}

@@

@@

\begin{itemize}
\item Networks on the chemistry side tend to be more for visualization
\item In many cases the network does not necessarily lead to nove output
\item A key utility of network approaches is the ability to integrate
  different data types such as small molecules and proteins
\item Networks represent an approach to representing multiple
  relationships in a single construct - multiple properties of a
  relationship between two entities can be encoded by edge properties
  (width, color, length). Similarly multiple properties of a given
  entity can be encoded in node properties (size, shape, color) 
\end{itemize}
\bibliography{paper}

>>>>

\newcommand{\rcdk}{\texttt{rcdk}\ }
\newcommand{\igraph}{\texttt{igraph}\ }

@@

\chapter{Exploring the Role of Small Molecules in Biological Systems Using Network Approaches}

@@

@@

\section{R as a Platform for Network Analyses in Drug Discovery}
\label{sec:r-as-platform}

@@

For a detailed review of \rcdk functionality, the reader is referred
to Guha \cite{Guha:2007aa}. For now we focus on simple network
analyses that involve chemical structures. Given a set of compounds
represented as SMILES strings, \rcdk can be used to parse them,
compute binary fingerprints and then compute a pairwise Tanimoto
similarity matrix, as shown below
\begin{lstlisting}
library(rcdk)
library(fingerprint)

mols <- parse.smiles(mipe$SMILES_STD)
fps <- lapply(mols, get.fingerprint, type='extended')
fpsim <- fp.sim.matrix(fps)
colnames(fpsim) <- mipe$SAMPLE_NAME
rownames(fpsim) <- mipe$SAMPLE_NAME
fpsim[ fpsim < 0.6 ] <- 0
\end{lstlisting}
In this code snippet, we compute a pairwise similarity matrix that we
will eventually use as an adjacency matrix to construct the similarity
network. We only keep those similarity values greater than 0.6 so as
to focus only on similar molecules.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.75\linewidth]{img/sim-network-connected}
  \caption{An example of a similarity network, where nodes are
    molecules and are connected if they exhibit a Tanimoto similarity
    greater than 0.6. Nodes are colored based on the PANTHER
    \cite{Mi:2005qq} class of their primary target.}
  \label{fig:simnet}
\end{figure}

The resultant similarity matrix can be treated as an adjacency matrix
and used as input to \igraph and subsequently plotted
\begin{lstlisting}
g <- graph.adjacency(fpsim, mode='undirected', 
                     weight=TRUE, diag=FALSE)
plot(g, layout=layout.fruchterman.reingold)
\end{lstlisting}
As example of such an analysis is shown in Figure \ref{fig:simnet}. In
this network, we consider a set of 1,341 compounds annotated with a
primary gene target. For each target, the highest level PANTHER
\cite{Mi:2005qq} classification was determined and used to color the
nodes. In addition, we ignore compounds that were less than 0.6
similar to any other compound (i.e., singleton nodes).

This is purely a visualization of a similarity matrix,
modified to focus on similar (according to a specified threshold)
compounds. Importantly, we have been to overlay a second type of
information (i.e., target class) on top of the similarity values and
is thus an example of the integrative nature of a network
representation.

Given the compound network and the associated target information, one
could easily identify protein-protein interactions (BioGRID
\cite{Chatr-Aryamontri:2015yf}, STRING \cite{Franceschini:2013qa} and
other sources) and generate a bipartite graph of compounds and protein
targets. Even in the trivial, annotated network show in Figure
\ref{fig:simnet} it is evident that chemical similarity is related to
functional similarity of the compounds' primary target - compounds
targeting cytoskeletal proteins tend to be similar to each other;
though interestingly, within the set of compounds targeting
cytoskeletal proteins there appear to be five distinct groupings. When
coupled to an interactive platform, this visualization can support
rapid identification and exploration of novel associations.

@@

In this chapter we have attempted to provide an overview of how
networks have been employed to analyze and visualize chemical
data. The fact that chemical structures themselves are graphs allows
one to apply a number of techniques from graph theory to characterize
chemical structure (such as the many graph invariants). However, with
the advent of large collections of chemical structures (from databases
such as Pubchem or high throughput screening programs) along with
associated data such as bioactivity and target annotations, a number
of workers have attempted to analyze the properties of such
collections using network techniques.

It is interesting to note that many applications of network methods to
chemical collections focus on visualization of the chemical data,
rather than using the network structure to draw insight into the
underlying chemical of biological phenomena. This is not to say that
network visualizations of chemical data are lacking. In fact, such
visualizations allow one to encode multiple properties of objects
(i.e., nodes) and relationships (i.e., edges) in a single visual form.

However, a key utility of network approaches is the ability to
integrate different data types (such as small molecules and their
protein targets) in to a single data structure. Drug-target or
drug-disease (or even drug-disease-target) networks are examples of
such approaches. Hartsperger et al \cite{Hartsperger:2010yg} describe
an approach to fuzzy clustering that is capable of identifying
subgraphs with heterogenous networks. A characteristic feature of such
$k$-partite networks is that it allows one to model the effects of
each class of entity on other classes of entities. This is exemplified
by random walk approaches on bipartite networks such as that described
by Chen et al \cite{Chen:2012qy} whose proposed a random walk approach
to predict drug-target interactions.

\bibliography{paper}
